{fairmetrics}: An R package for group fairness evaluation

### 2025-06-02 paper.Rmd
Edit summary.

Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for particular groups, particularly those defined by protected attributes such as race, gender, or age. Evaluating fairness is a critical aspect of ML model development, as biased models can perpetuate structural inequalities.  The {fairmetrics} R package offers a user-friendly framework for rigorously evaluating numerous group-based fairness criteria, including metrics based on independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity). Group-based fairness criteria assess whether a model is equally accurate or well-calibrated across a set of predefined groups so that appropriate bias mitigation strategies can be implemented. {fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions and includes an example dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].

@@

# Evaluating Fairness Criteria

<!-- The figure looks good but do not use any abbreviations (i.e., Predicted probabilities).  Also, we use "protected" not sensitive attribute in paper so needs an update.  Also add negative predictive parity once added to the package. For combined: What's the difference between get_all and get_fairness? This isn't clear to me. Also I think you should have an argument for level for the CIs - that is, users input 0.95 for 95\% or 0.99 etc as that should be adjustable.

BEN RESPONSE:

1. Updated the visual to not use abbreviations.
2. Changed "sensitive' attribute to "protected" attribute
3. Added negative predictive parity to package (`eval_neg_pred_parity()`; renamed `eval_pred_parity()` to `eval_pos_pred_parity()`). Submitted & Uploaded to CRAN. Added to the visual as well.
4.  Noted. Refactored the code to allow get_fairness_metrics() to cover the functionality of get_all_metrics(). This can be specified with the confint = FALSE.
5.  There is an argument level for the CIs, its the `alpha` parameter which by default is set to 0.05. Added the argument to the code. Do you want me to talk about it in the text as well? 
 -->
The input to the {fairmetrics} package is a data frame or tibble containing the model's predicted probabilities, the true outcomes, and the protected attribute of interest. \hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}. Users can evaluate a model for a specific criterion or multiple group fairness criteria using the combined metrics function.

>>>>

Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for specific groups, particularly those defined by protected attributes such as race, gender, or age. Evaluating fairness is a critical aspect of ML model development, as biased models can perpetuate structural inequalities.  The {fairmetrics} R package offers a user-friendly framework for rigorously evaluating numerous group-based fairness criteria, including metrics based on independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity). Group-based fairness criteria assess whether a model is equally accurate or well-calibrated across a set of predefined groups so that appropriate bias mitigation strategies can be implemented. {fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions and includes an example dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].

@@

# Evaluating Fairness Criteria
The input to the {fairmetrics} package is a data frame or tibble containing the model's predicted probabilities, the true outcomes, and the protected attribute of interest. \hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}. Users can evaluate a model for a specific criterion or multiple group fairness criteria using the combined metrics function.
